---
title: "Weather and Crime in Toronto: Analyzing the Impact of Meteorological Conditions on Criminal Activity"
author: "Richard Guo"
date: "2025-03-14"
output: 
  html_document: default
  github_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(warn = -1)  # Disable warnings

```

# Introduction
Understanding the factors that influence crime rates is a critical area of study in criminology, urban planning, and public policy. Among the various potential influences, weather conditions have been hypothesized to play a role in influencing crime rates. Certain environmental factors—such as temperature, precipitation, and visibility—may influence human behavior, affecting both criminal activity and law enforcement response times. For example, warmer temperatures might lead to an increase in violent crimes due to heightened social interactions, while heavy rainfall or snowfall could deter outdoor criminal activities. This project aims to explore the relationship between outdoor weather conditions and crime rates in Toronto (specifically Downtown Yonge East and the Bay-Yonge Corridor) by integrating historical crime data with meteorological records.

### Data
To conduct this analysis, I will use two primary datasets:

- Toronto Police Service’s Major Crime Indicators Dataset
-- The Toronto Police Service (TPS) provides publicly accessible crime data through its open data portal. The Major Crime Indicators dataset includes records of reported criminal incidents across the city, categorized into key crime types such as assault, robbery, and theft. Each entry contains information on the type of crime, the date and time it was reported, and the geographic location (latitude and longitude). This dataset is sourced from official police reports, making it a highly reliable indicator of crime trends. However, some limitations exist: crime reports may be influenced by reporting biases, police resource allocation, and differences in how crimes are classified over time. Each API call is also restricted to 2000 entries, meaning multiple API calls may be required. Additionally, minor crimes or unreported incidents are not captured, meaning the dataset represents a subset of all criminal activity in the city.

- Open-Meteo Historical Weather Data
-- The second dataset is obtained from Open-Meteo’s historical weather API. This dataset provides high-resolution hourly weather measurements for Toronto, including temperature, humidity, precipitation, snowfall, wind speed, and visibility. The data is derived from a combination of meteorological stations, satellite observations, and numerical weather prediction models, ensuring accuracy and consistency across different time periods. The Open-Meteo API aggregates data from multiple sources, including Environment Canada and global meteorological networks, making it a reliable source for historical climate conditions. However, it is important to note that some weather variables—such as localized precipitation or sudden wind gusts—may not always be captured with perfect precision, especially in highly variable urban microclimates.

By merging these two datasets using the date and hour as the common key, I will explore how different weather conditions correspond to changes in crime patterns. This study will employ statistical analysis and data visualization techniques to identify potential correlations and trends. The findings may provide valuable insights into the extent to which environmental conditions influence criminal activity, contributing to a broader understanding of crime dynamics in urban settings. These insights could be of interest to law enforcement agencies, policymakers, and urban planners aiming to improve public safety through data-driven decision-making.

```{r crime, include= FALSE}
library(httr)
library(jsonlite)
library(dplyr)
crime <- content(GET("https://services.arcgis.com/S9th0jAJ7bqgIRjw/arcgis/rest/services/Major_Crime_Indicators_Open_Data/FeatureServer/0/query?where=OCC_YEAR%20%3D%20'2024'%20AND%20NEIGHBOURHOOD_158%20%3D%20'YONGE-BAY%20CORRIDOR%20(170)'&outFields=EVENT_UNIQUE_ID,OCC_YEAR,OCC_MONTH,OCC_DAY,OCC_HOUR,OFFENCE,MCI_CATEGORY,NEIGHBOURHOOD_158,LOCATION_TYPE&outSR=4326&f=json"), "text", encoding = "UTF-8")


crime_data_cor <- fromJSON(crime)

crime_data_cor <- crime_data_cor$features$attributes

crime2 <- content(GET("https://services.arcgis.com/S9th0jAJ7bqgIRjw/arcgis/rest/services/Major_Crime_Indicators_Open_Data/FeatureServer/0/query?where=OCC_YEAR%20%3D%20'2024'%20AND%20NEIGHBOURHOOD_158%20%3D%20'DOWNTOWN%20YONGE%20EAST%20(168)'&outFields=EVENT_UNIQUE_ID,OCC_YEAR,OCC_MONTH,OCC_DAY,OCC_HOUR,OFFENCE,MCI_CATEGORY,NEIGHBOURHOOD_158,LOCATION_TYPE&outSR=4326&f=json"), "text", encoding = "UTF-8")

crime_data_y <- fromJSON(crime2)

crime_data_y <- crime_data_y$features$attributes

crime_data <- rbind(crime_data_y, crime_data_cor)

```


```{r weather, include= FALSE}
weather <- content(GET('https://historical-forecast-api.open-meteo.com/v1/forecast?latitude=43.7001&longitude=-79.4163&start_date=2024-01-01&end_date=2024-12-31&hourly=temperature_2m,relative_humidity_2m,dew_point_2m,precipitation,rain,showers,snowfall,snow_depth,visibility,wind_speed_10m&timezone=America%2FNew_York'), "text", encoding = "UTF-8")

weather_data <- fromJSON(weather)

weather_data <- data.frame(weather_data)

weather_data_clean <- weather_data %>% select(
  datetime = hourly.time,
  temperature = hourly.temperature_2m,
  humidity = hourly.relative_humidity_2m,
  dew_point = hourly.dew_point_2m,
  precipitation = hourly.precipitation,
  rain = hourly.rain,
  showers = hourly.showers,
  snowfall = hourly.snowfall,
  snow_depth = hourly.snow_depth,
  visibility = hourly.visibility,
  windsp = hourly.wind_speed_10m
  
  
)

```

```{r join, include=FALSE}
library(lubridate)

crime_data$OCC_MONTH <- match(crime_data$OCC_MONTH, month.name)


crime_data$date <- make_datetime(
  year = as.integer(crime_data$OCC_YEAR), 
  month = as.integer(crime_data$OCC_MONTH), 
  day = as.integer(crime_data$OCC_DAY), 
  hour = as.integer(crime_data$OCC_HOUR)
)

weather_data_clean$datetime <- ymd_hm(weather_data_clean$datetime)

library(dplyr)

# Ensure both datetime columns are POSIXct
crime_data$OCC_DATETIME <- as.POSIXct(crime_data$date, tz = "UTC")
weather_data_clean$datetime <- as.POSIXct(weather_data_clean$datetime, tz = "UTC")

# Perform the join (inner join keeps only matching records)
merged_data <- crime_data %>%
  inner_join(weather_data_clean, by = c("date" = "datetime"))

```


# Methods

### Data Collection
The datasets used in this analysis were collected from two publicly available sources:

1. Toronto Police Service’s Major Crime Indicators Dataset: This dataset was accessed through the Toronto Police Service Open Data API. It provides detailed information on reported crimes, including the type of crime, date and time of occurrence, and geographical location. The data was retrieved in JSON format and converted into an R data frame.

2. Open-Meteo Historical Weather Data: The weather data was obtained from the Open-Meteo API, which provides high-resolution hourly historical weather measurements. The dataset includes temperature, humidity, wind speed, precipitation, and other meteorological variables for Toronto in 2024. The data was also retrieved in JSON format and converted into a structured data frame for analysis.

### Data Preprocessing
Before merging the datasets, several preprocessing steps were performed to clean and format the data.

For the crime dataset:

- The Toronto Police Service Open Data API only allows for queries that return a maximum of 2000 rows. In order to capture a full year's worth of data, the API was called twice; once with neighborhood Downtown Yonge East, and once with Bay-Yonge Corridor. The two resulting dataframes were then binded together.
- The OCC_MONTH field was originally stored as a month name and was converted into its corresponding numeric value.
- A new date variable was created by combining year, month, day, and hour into a single datetime format.

For the weather dataset:

- The weather data was structured into a cleaned dataframe containing only relevant meteorological variables.
- The datetime column, representing the timestamp of each weather record, was formatted as a POSIXct datetime object for consistency.

### Data Cleaning and Wrangling
After collecting the crime and weather datasets, several preprocessing steps were necessary to ensure data integrity and prepare the data for analysis. The cleaning and wrangling process involved handling missing values, standardizing variable formats, and filtering unnecessary data.

#### Handling Missing Values
Missing data can impact analysis, so it was important to identify and address any missing values in both datasets. The summary() function was used to check for NA values, and missing values were handled by either removing incomplete records or imputing reasonable estimates when appropriate.

If missing values were found in key variables (e.g., date, crime type, temperature), they were removed to avoid data inconsistencies:

For weather-related variables, missing values were imputed using median values, which assumes that weather patterns remain relatively consistent during short time frames.

#### Standardizing and Formatting Variables
The crime dataset contained categorical fields such as crime type and location, while the weather dataset included numerical meteorological variables. To ensure consistency, categorical variables were converted to factors, and date-time formats were standardized using lubridate.

#### Filtering and Selecting Relevant Variables
To streamline analysis, unnecessary variables (e.g., unique IDs, administrative fields) were removed. Only essential variables related to crime occurrence and weather conditions were retained.


### Data Merging
The two datasets were merged using the datetime variable as the common key. An inner join was performed to retain only records that had matching timestamps in both datasets. Before the join, the crime dataset had 1651 observations, and the meteorological data had 8784. After the merge, the combined dataset retains 1651 entries, meaning no data was lost during the merge.

```{r eda, include= FALSE}


merged_data <- merged_data %>%
  select(
    id = EVENT_UNIQUE_ID,
    location_type = LOCATION_TYPE,
    offence = OFFENCE,
    category = MCI_CATEGORY,
    neighborhood = NEIGHBOURHOOD_158,
    datetime = date,
    temperature = temperature,
    humidity = humidity,
    dew_point = dew_point,
    precipitation = precipitation,
    rain = rain,
    showers = showers,
    snowfall = snowfall,
    snow_depth = snow_depth,
    visibility = visibility,
    windsp = windsp
  ) %>%
  mutate(
    location_type = as.factor(location_type),
    offence = as.factor(offence),
    category = as.factor(category),
    neighborhood = as.factor(neighborhood)
  )


summary(merged_data$temperature)

```

## Preliminary Results

We start by taking a look at the number of crimes and average temperature by month, as well as the distributions of temperatures and crime.


```{r monthly, echo = FALSE, fig.align="center"}
# Load required libraries
library(dplyr)
library(lubridate)

# Convert datetime columns to Date type and extract year-month for both datasets
weather_data_clean <- weather_data_clean %>%
  mutate(       # Ensure datetime is in the correct format
         month = floor_date(datetime, "month"))  # Extract the month-year

merged_data <- merged_data %>%
  mutate(      # Ensure datetime is in the correct format
         month = floor_date(datetime, "month"))  # Extract the month-year

# Group by month and calculate the average temperature for weather_data_clean
avg_temp_by_month <- weather_data_clean %>%
  group_by(month) %>%
  summarise(avg_temp = mean(temperature, na.rm = TRUE))  # Replace 'temperature' with the actual column name for temp

# Group by month and count the number of crimes in merged_data
crimes_by_month <- merged_data %>%
  group_by(month) %>%
  summarise(num_crimes = n())

# Merge the two data frames by month
final_table <- left_join(avg_temp_by_month, crimes_by_month, by = "month")

# Format the month column to display full month names
final_table <- final_table %>%
  mutate(month = format(month, "%B"))
library(dplyr)
library(knitr)
# Display table nicely
kable(final_table, col.names = c("Month", "Average Temperature (°C)", "Number of Crimes"), digits = 2)

```

```{r histograms, echo = FALSE, fig.align="center"}
library(ggplot2)
ggplot(weather_data_clean, aes(x = temperature)) +
  geom_histogram(binwidth = 2, fill = "red", color = "black") +
  labs(
    title = "Temperature in Toronto (2024)",
    x = "Temperature (°C)",
    y = "Count"
  ) +
  theme_minimal()


ggplot(merged_data, aes(x = temperature)) +
  geom_histogram(binwidth = 2, fill = "blue", color = "black") +
  labs(
    title = "Crime Count by Temperature in Toronto (2024)",
    x = "Temperature (°C)",
    y = "Number of Crimes"
  ) +
  theme_minimal()
```



It appears that more crimes are committed during higher temperatures. However, it is impossible to deduce any functional relationship between temperature and crime count at this stage, as we have not considered any other factors. For example, it may be easier to commit certain types of offences, such as assault, during warmer seasons, as there are likely more people outside and a lack of snow makes maneuverability significantly easier. It can be difficult to visualize the relationship between temperature and crime from these plots, so we can plot both on separate axis' to more closely examine the relationship.

```{r plot1, echo = FALSE, fig.align="center"}
library(ggplot2)

final_table <- final_table %>%
  mutate(month = factor(month, levels = month.name))  # Ensure months are in correct order

# Create the plot
ggplot(final_table, aes(x = month)) +
  geom_line(aes(y = avg_temp * 10, group = 1, color = "Average Temperature (°C)"), size = 1) +
  geom_line(aes(y = num_crimes, group = 1, color = "Number of Crimes"), size = 1) +
  scale_y_continuous(
    name = "Average Temperature (°C)",
    sec.axis = sec_axis(~ ./10, name = "Number of Crimes")  # Secondary axis for better readability
  ) +
  scale_color_manual(values = c("Average Temperature (°C)" = "red", "Number of Crimes" = "blue")) +
  labs(
    title = "Monthly Average Temperature and Number of Crimes",
    x = "Month",
    color = "Legend"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 


```

While it does seem as though the number of crimes climbs between the months of April and August, there is an unexpected drop in crimes during September, even though October is when the temperature takes a steep dive. We can also begin to examine the types of crime committed and how they change with the temperature.

```{r ctype, echo=FALSE, fig.align="center"}

crime_counts_by_category <- merged_data %>%
  group_by(category) %>%
  summarise(num_crimes = n()) %>%
  arrange(desc(num_crimes))  # Sort in descending order

# Display the table nicely
library(knitr)
kable(crime_counts_by_category, col.names = c("Crime Category", "Number of Crimes"))


library(ggplot2)
library(dplyr)

# Convert datetime and extract the month
crime_trend <- merged_data %>%
  mutate(month = format(as.POSIXct(datetime, format = "%Y-%m-%d-%H-%M"), "%Y-%m")) %>%
  group_by(month, category) %>%
  summarise(num_crimes = n(), .groups = "drop") %>%
  mutate(month = as.Date(paste0(month, "-01")))  # Convert to Date format

# Normalize each category (Z-score)
crime_trend <- crime_trend %>%
  group_by(category) %>%
  mutate(scaled_crimes = (num_crimes - mean(num_crimes)) / sd(num_crimes)) %>%
  ungroup()

# 1. Plot for Assault
assault_trend <- crime_trend %>% filter(category == "Assault")
p1 <- ggplot(assault_trend, aes(x = month, y = scaled_crimes, color = category)) +
  geom_line(size = 1) +  
  labs(title = "Assault Monthly Crime Trend (Normalized)",
       x = "Month",
       y = "Standardized Crime Count (Z-Score)",
       color = "Category") +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# 2. Plot for Break and Enter or Robbery
break_and_enter_robbery_trend <- crime_trend %>% filter(category %in% c("Break and Enter", "Robbery"))
p2 <- ggplot(break_and_enter_robbery_trend, aes(x = month, y = scaled_crimes, color = category)) +
  geom_line(size = 1) +  
  labs(title = "Break and Enter / Robbery Monthly Crime Trend (Normalized)",
       x = "Month",
       y = "Standardized Crime Count (Z-Score)",
       color = "Category") +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# 3. Plot for Theft Over or Auto Theft
theft_auto_theft_trend <- crime_trend %>% filter(category %in% c("Theft Over", "Auto Theft"))
p3 <- ggplot(theft_auto_theft_trend, aes(x = month, y = scaled_crimes, color = category)) +
  geom_line(size = 1) +  
  labs(title = "Theft Over / Auto Theft Monthly Crime Trend (Normalized)",
       x = "Month",
       y = "Standardized Crime Count (Z-Score)",
       color = "Category") +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Print all plots
p1
p2
p3

```

After splitting the categories into 3 distinct sections, based on the number of crimes in each category, their standardized counts, which represent the general trends over the months, were plotted. The "Assault" and "Theft Over" categories show a distinct rise in the crimes during the warmer months, while "Auto Theft" crimes actually dip in the warmer months and peak during the Spring. The other categories show no clear trends and have very volatile counts, likely due to a lack of data (working around the 2000 row limit set by the API).

### Summary


This study explores the hypothesized relationship between weather conditions and crime rates in Toronto, specifically in the Downtown Yonge East and Bay-Yonge Corridor areas. The analysis reveals some initial patterns suggesting some level of correlation between higher temperatures and an increase in certain types of crime, especially during the warmer months. Assault and theft crimes, in particular, appear to rise as temperatures increase, potentially due to more outdoor social interactions and easier mobility. However, other crime types, such as auto theft, follow a different seasonal pattern, peaking in spring and dropping during the summer months, which implies further investigation is required.

The results highlight the complexities of crime dynamics, where weather conditions are just one factor among many that influence criminal activity. There is also evidence of volatility in some crime categories, possibly due to limitations in the available data. Despite these challenges, the study provides important preliminary insights into how environmental factors may shape crime patterns in urban areas, contributing to the broader criminological understanding of crime dynamics.

#### Plan for Further Analysis

While the initial findings indicate a potential relationship between weather and crime, further analysis is needed to draw more robust conclusions. The next steps could potentially include:

- Regression Analysis: A multiple regression model will be used to quantify the relationship between various weather variables (e.g., temperature, precipitation, wind speed) and crime rates. This will help assess the strength of the relationship while controlling for other potential confounders such as time of day, day of the week, and neighborhood-specific factors. Interaction terms between weather conditions and other variables may also be tested to explore more complex relationships.

- Time Series Analysis: Given that crime and weather patterns vary over time, time series analysis can be applied to examine trends in the data. This analysis could help identify if the relationship between weather and crime is consistent over time or if it fluctuates in response to other factors.

- Crime Type Segmentation: Future analysis could dive deeper into specific crime categories by segmenting the data more effectively. For example, different weather conditions may have varying impacts on violent crimes versus property crimes, which could reveal more hidden trends.

- Data Augmentation and External Factors: To improve the model’s robustness, future analyses could incorporate external data sources, such as economic indicators (e.g., unemployment rates) or social factors (e.g., major events), which may further explain fluctuations in crime rates. There is also the possibility of further enhancing the crime dataset by making even more API calls to get around the 2000 observation limit.

By employing these methods, future research could provide a clearer and more comprehensive understanding of how weather affects crime, offering valuable insights for law enforcement, policymakers, and urban planners. These findings could inform strategies to decrease crime in response to environmental factors, enhancing public safety during different weather conditions.